The \dealii Library, Version 9.5

### 2023-06-29 9.5/paper.tex
Diverse

The major changes of this release are:
%
\begin{itemize}
  \item Substantial updates and extensions to \dealii{}'s interfaces
    to other libraries (see Section~\ref{sec:external}). This
    includes, in particular, the integration of \kokkos (Section~\ref{sec:kokkos});
    additions and updates to the \petsc and \trilinos
    interfaces
    (Sections~\ref{sec:petsc} and \ref{sec:trilinos}).
  \item Uniform handling of nonlinear solver
  packages (Section~\ref{sec:nonlinear}) and a uniform
    way of defining callbacks used by external libraries (Section~\ref{sec:callbacks}).
  \item Advances in matrix-free infrastructure (see Section~\ref{sec:mf}).
  \item Advances in non-matching support (see Section~\ref{sec:nonmatching}).
  \item New features related to linear algebra (see Section~\ref{sec:lac}).
  \item C++ language modernization (see Section~\ref{sec:language}).
\end{itemize}
%

@@

In the current release, \texttt{LinearAlgebra::distributed::Vector} and the 
\texttt{CUDAWrappers::MatrixFree} framework are using \kokkos. This allows them 
to work on all the architectures supported by \kokkos. The \kokkos backend used 
by \dealii is \texttt{Kokkos::DefaultExecutionSpace} which corresponds to the 
highest available backend in the hierarchy device, host-parallel, and host-serial 
at the time Kokkos was configured.

@@

Vector and matrix classes of the \petsc wrappers have been extended with an additional constructor
that can wrap an already existing \petsc vector or matrix, respectively.
The \texttt{BlockVector} and
\texttt{BlockSparseMatrix} classes now internally use
\petsc nested objects, i.e. \texttt{VECNEST} and  \texttt{MATNEST} respectively.
We have added a new class \texttt{PETScWrappers::PreconditionShell} to support user-defined
preconditioning that can be as simply customized as the following code
snippet:
\begin{c++}
PETScWrappers::PreconditionShell preconditioner(/*...*/);

@@

The \petsc~\snes sub-package solves systems of nonlinear equations of the form $F(x) = 0$.
The interface to \snes has been modeled on the already existing interface
to the \kinsol solver from the \sundials package, and the nonlinear problem
is specified with a set of callbacks that are used by the solver:
\begin{c++}
PETScWrappers::NonlinearSolver<VectorType> solver(/*...*/);

solver.residual            = [&](const VectorType &x,
                                 VectorType       &F) {/*...*/};
solver.setup_jacobian      = [&](const VectorType &x) {/*...*/};
solver.solve_with_jacobian = [&](const VectorType &rhs,
                                 VectorType       &sol) {/*...*/};
solver.solve(x);
\end{c++}
The default configuration is set up to use a Jacobian-free Newton-Krylov (JFNK)
approach~\cite{knoll2004jacobian}, using \texttt{solve\_with\_jacobian} as linear solver.
Numerous other solver configurations are possible and can be selected
programmatically using the constructor arguments,
or via the powerful command line customization of \petsc.
This includes, for example, Quasi-Newton methods, Anderson's acceleration, and
nonlinear preconditioning~\cite{brune2015composing}. The step-77
tutorial program has gained a section that discusses the use of these
new \snes interfaces.

@@

Secondly, we have added a wrapper to \texttt{NOX}, a
nonlinear solver library that is similar to both the \kinsol{} solver from
the \sundials{} package to which we already had interfaces, and also
to the \snes{} collections of functions from \petsc (see also
Section~\ref{sec:petsc}). The interfaces to these three solvers are
essentially the same and require setting function objects for
computing the residual of the nonlinear problem, along with setting
up, applying, and solving with the Jacobian matrix:
\begin{c++}
TrilinosWrappers::NOXSolver<VectorType> solver(/*...*/);

solver.residual            = [ ](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
solver.apply_jacobian      = [&](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.solve_with_jacobian = [&](const VectorType &src,
                                 VectorType       &dst,
                                 const double      tol) {/*...*/};

solver.solve(solution);
\end{c++}

We refer interested readers to our documentation for other more advanced
functions of this wrapper which are related to the differences in features of these libraries.
In particular, we have added the possibility to reuse the preconditioner between
nonlinear steps (also known as \textit{preconditioner lagging}), which is
unfortunately natively only  supported in the official \texttt{EPetra} implementations.

@@

With the current release, \dealii{} now supports solvers for nonlinear problems
provided by several different external packages:
(i) \kinsol{} (part of \sundials{}); (ii) \snes{} (part of \petsc{}, see Section~\ref{sec:petsc}); and (iii) \texttt{NOX} (part of \trilinos{}, see Section~\ref{sec:trilinos}).
The wrappers are provided by the classes \texttt{PETScWrappers::NonlinearSolver<VectorType>}, \texttt{SUNDIALS::KINSOL<VectorType>}, and \texttt{TrilinosWrappers::NOXSolver<VectorType>}, respectively.

@@

Here is a complete example of an MPI parallel Newton solver that
automatically chooses one of the available packages:
\todo[inline]{The NOX example was nicely indented. Can we do the same
  here? Perhaps if we call the variable \texttt{nl\_solver}?}
\todo[inline]{But separately: Does the code snippet really provide any
  additional value? It is for all practical purposes identical to the
  NOX one, and the NOX one is already identical to the SNES one. Do we
need to repeat things a third time, or could one just use one or two
sentences to express the same?}
\begin{c++}
    using NLSolver = NonlinearSolverSelector<LA::MPI::Vector>;

    NLSolver::AdditionalData additional_data(NLSolver::automatic,
                                             NLSolver::newton);
    NLSolver nonlinear_solver(additional_data, mpi_communicator);

    nonlinear_solver.reinit_vector = [&](LA::MPI::Vector &x)
                                     {/*...*/};
    nonlinear_solver.residual = [&](const LA::MPI::Vector &evaluation_point,
                                    LA::MPI::Vector &      residual)
                                {/*...*/};
    nonlinear_solver.setup_jacobian = [&](const LA::MPI::Vector &current_u)
                                      {/*...*/};
    nonlinear_solver.solve_with_jacobian = [&](const LA::MPI::Vector &rhs,
                                               LA::MPI::Vector &      dst,
                                               const double tolerance)
                                           {/*...*/};

    nonlinear_solver.solve(current_solution);
\end{c++}

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/mesh_0.png}
  \end{subfigure}\qquad
  \hfill
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/mesh_1.png}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/mesh_2.png}
  \end{subfigure}
  \caption{\it Example of non-nested multigrid levels.}\label{fig:nonnested}
\end{figure}

@@

two_level_transfer.reinit(dof_handler_fine, dof_handler_coarse,
                          // the following parameters are optional
                          mapping_fine, mapping_coarse,
                          constraints_fine, constraints_coarse)
\end{c++}
The resulting object can be passed to \texttt{MGTransferGlobalCoarsening} (and
\texttt{MGTransferBlockGlobalCoarsening}) in the usual way. The
interface of the new class is purposefully similar to
\texttt{MGTwoLevelTransfer},
with the difference that the former (i) also
takes \texttt{Mapping} instances, and (ii) is not limited to the case
that the coarser of the two \texttt{DoFHandler} instances can be generated by a coarsening step from the fine
mesh.

At the time of writing, the \texttt{MGTwoLevelTransferNonNested} operator provides a prolongation operator $\mathcal{P}^{(f,c)}$
that prolongates a vector $x^{(c)}$ from the coarse space to the fine space:
\begin{align*}
x^{(f)} = \mathcal{P}^{(f, c)}  x^{(c)},
\end{align*}
by a pointwise interpolation (injection) from the coarse mesh to the support points of the fine mesh. This is done
by looping over all coarse cells and interpolating to all (fine support) points that fall into the cell with
\texttt{FEPointEvaluation}. The local result is scattered into $x^{(f)}$, which is finalized by a
communication and a weighting step that is necessary as multiple elements could add to the same global entry.

The current implementation works for scalar and vectorial, continuous (\texttt{FE\_Q})
and discontinuous (\texttt{FE\_DGQ}) elements on hypercube-shaped cells. We plan to support
simplex elements and to add support
for projection, in addition to the pointwise interpolation outlined
above. Furthermore, we envision providing utility tools to generate
coarser
meshes, based on a fine mesh. Currently, the users themselves
have to provide such meshes,
e.g., by generating meshes with different cell sizes in external mesh-generation tools.

@@

\todo[inline]{Matthias: Write}

@@

The 9.5 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_4_2_and_9_5_0.html}
     {around XXX incompatible changes};
\todo{Fix number before release}
see \cite{changes95}. Many of these
incompatibilities change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item
  The class \texttt{hp::DoFHandler} has been removed. The
  \texttt{DoFHandler} class now implements all $hp$-related
  functionality, as it has for the past several releases.

@@

  \item
  This rework also affects the serialization process of active FE
  indices. You will need to recreate your serialized data \textit{if
    and only if} you work in $hp$-mode, but can continue to use
  previously generated data otherwise. For
  \texttt{parallel::distributed::Triangulation} when used with non-$hp$ data, it is sufficient to increase the version in the metadata file from ``4'' to ``5''.

\item Several old interfaces to \texttt{MatrixFree} have been removed, e.g.,
  initialization functions without \texttt{Mapping} argument, the query
  to the number of cell batches, \texttt{DoFHandler} objects. In
  each case, new interfaces are available.
\todo[inline]{The sentence part ``e.g., initialization
  functions...objects'' does not sound grammatically correct. Should
  it have been ``and the functions querying \texttt{DoFHandler}
  objects''? Please fix and remove this box.}
\end{itemize}

>>>>

The major changes of this release are:
%
\begin{itemize}
  \item Substantial updates and extensions to \dealii{}'s interfaces
    to other libraries (see Section~\ref{sec:external}). This
    includes, in particular, the integration of \kokkos (Section~\ref{sec:kokkos});
    additions and updates to the \petsc and \trilinos
    interfaces
    (Sections~\ref{sec:petsc} and \ref{sec:trilinos}, respectively).
  \item Uniform handling of nonlinear solver
  packages (Section~\ref{sec:nonlinear}) and a uniform
    way of defining callbacks used by external libraries (Section~\ref{sec:callbacks}).
  \item Advances in matrix-free infrastructure (see Section~\ref{sec:mf}).
  \item Advances in non-matching support (see Section~\ref{sec:nonmatching}).
  \item New features related to linear algebra (see Section~\ref{sec:lac}).
  \item C++ language modernization (see Section~\ref{sec:language}).
\end{itemize}
%

@@

In the current release, \texttt{LinearAlgebra::distributed::Vector} and the 
\texttt{CUDAWrappers::Matrix\allowbreak{}Free} framework are using \kokkos. This allows them 
to work on all the architectures supported by \kokkos. The \kokkos backend used 
by \dealii is \texttt{Kokkos::DefaultExecutionSpace} which corresponds to the 
highest available backend in the hierarchy device, host-parallel, and host-serial 
at the time Kokkos was configured.

@@

Vector and matrix classes of the \petsc wrappers have been extended with an additional constructor
that can wrap an already existing \petsc vector or matrix, respectively.
The \texttt{BlockVector} and
\texttt{BlockSparseMatrix} classes now internally use
\petsc nested objects, i.e. \texttt{VECNEST} and  \texttt{MATNEST} respectively.
We have added a new class \texttt{PETScWrappers::PreconditionShell} to support user-defined
preconditioning that can be simply customized as the following code
snippet shows:
\begin{c++}
PETScWrappers::PreconditionShell preconditioner(/*...*/);

@@

The \petsc~\snes sub-package solves systems of nonlinear equations of the form $F(x) = 0$.
The interface to \snes has been modeled on the already existing interface
to the \kinsol solver from the \sundials package, and the nonlinear problem
is specified with a set of callbacks that are used by the solver.
%\begin{c++}
%PETScWrappers::NonlinearSolver<VectorType> solver(/*...*/);
%
%solver.residual            = [&](const VectorType &x,
%                                 VectorType       &F) {/*...*/};
%solver.setup_jacobian      = [&](const VectorType &x) {/*...*/};
%solver.solve_with_jacobian = [&](const VectorType &rhs,
%                                 VectorType       &sol) {/*...*/};
%solver.solve(x);
%\end{c++}
The default configuration is set up to use a Jacobian-free Newton-Krylov (JFNK)
approach~\cite{knoll2004jacobian}, using \texttt{solve\_with\_jacobian} as linear solver.
Numerous other solver configurations are possible and can be selected
programmatically using the constructor arguments,
or via the powerful command line customization of \petsc.
This includes, for example, Quasi-Newton methods, Anderson's acceleration, and
nonlinear preconditioning~\cite{brune2015composing}. The step-77
tutorial program has gained a section that discusses the use of these
new \snes interfaces.

@@

Secondly, we have added a wrapper to \texttt{NOX}, a
nonlinear solver library that is similar to both the \kinsol{} solver from
the \sundials{} package to which we already had interfaces, and also
to the \snes{} collections of functions from \petsc (see also
Section~\ref{sec:petsc}). 
%The interfaces to these three solvers are
%essentially the same and require setting function objects for
%computing the residual of the nonlinear problem, along with setting
%up, applying, and solving with the Jacobian matrix:
%\begin{c++}
%TrilinosWrappers::NOXSolver<VectorType> solver(/*...*/);
%
%solver.residual            = [ ](const VectorType &src,
%                                 VectorType       &dst) {/*...*/};
%solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
%solver.apply_jacobian      = [&](const VectorType &src,
%                                 VectorType       &dst) {/*...*/};
%solver.solve_with_jacobian = [&](const VectorType &src,
%                                 VectorType       &dst,
%                                 const double      tol) {/*...*/};
%
%solver.solve(solution);
%\end{c++}
%
We refer interested readers to our documentation for other more advanced
functions of this wrapper which are related to the differences in features of these libraries.
In particular, we have added the possibility to reuse the preconditioner between
nonlinear steps (also known as \textit{preconditioner lagging}), which is
unfortunately natively only  supported in the official \texttt{EPetra} implementations.

@@

With the current release, \dealii{} now supports solvers for nonlinear problems
provided by several different external packages:
(i) \kinsol{} (part of \sundials{}); (ii) \snes{} (part of \petsc{}, see Section~\ref{sec:petsc}); and (iii) \texttt{NOX} (part of \trilinos{}, see Section~\ref{sec:trilinos}).
The wrappers are provided by the classes \texttt{PETScWrappers::NonlinearSolver\allowbreak{}<VectorType>}, \texttt{SUNDIALS::KINSOL<VectorType>}, and \texttt{TrilinosWrappers::NOXSolver<Vector\allowbreak{}Type>}, respectively.

@@

The following code snippet shows a complete example using the new class for applying a Newton solver
by automatically choosing one of the available packages:
\begin{c++}
using NLS = NonlinearSolverSelector<VectorType>;

NLS::AdditionalData additional_data(
  NLS::automatic, // other options: kinsol, nox, petsc_snes 
  NLS::newton);

NLS nonlinear_solver(additional_data, mpi_communicator);

solver.reinit_vector       = [ ](VectorType &x) {/*...*/};    
solver.residual            = [ ](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
solver.apply_jacobian      = [&](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.solve_with_jacobian = [&](const VectorType &src,
                                 VectorType       &dst,
                                 const double      tol) {/*...*/};

solver.solve(current_solution);    
\end{c++}

Not that while the selector class provides a unified interface and therefore allows to
easily switch between backends, there are cases when users need to use the underlying
wrapper classes, if, e.g., the functionality is only provided by one implementation.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/mesh_0.png}
  \end{subfigure}\qquad
  \hfill
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/mesh_1.png}
  \end{subfigure}
  \hfill
  \begin{subfigure}[b]{0.28\textwidth}
    \centering
    \includegraphics[width=\textwidth]{png/mesh_2.png}
  \end{subfigure}
  \caption{\it Example of non-nested multigrid levels.\todo[inline]{Make gap uniform!}}\label{fig:nonnested}
\end{figure}

@@

two_level_transfer.reinit(dof_handler_fine, dof_handler_coarse,
                          mapping_fine, mapping_coarse,
                          constraints_fine, constraints_coarse)
\end{c++}
The resulting object can be passed to \texttt{MGTransferGlobalCoarsening} (and
\texttt{MGTransferBlockGlobal\allowbreak{}Coarsening}) in the usual way. The
interface of the new class is purposefully similar to
\texttt{MGTwoLevel\allowbreak{}Transfer},
with the difference that the former (i) also
takes \texttt{Mapping} instances, and (ii) is not limited to the case
that the coarser of the two \texttt{DoFHandler} instances can be generated by a coarsening step from the fine
mesh.

At the time of writing, the \texttt{MGTwoLevelTransferNonNested} operator provides a prolongation operator
that prolongates a vector from the coarse space to the fine space,
by a pointwise interpolation (injection) from the coarse mesh to the support points of the fine mesh. This is done
by looping over all coarse cells and interpolating to all (fine support) points that fall into the cell with
\texttt{FEPointEvaluation}. The local result is scattered into the fine vector, which is finalized by a
communication and a weighting step that is necessary as multiple elements could add to the same global entry.

The current implementation works for scalar, continuous (\texttt{FE\_Q})
and discontinuous (\texttt{FE\_DGQ}) elements on hypercube-shaped cells. We plan to support
vectorial and simplex elements and to add support
for projection, in addition to the pointwise interpolation outlined
above. Furthermore, we envision providing utility tools to generate
coarser
meshes, based on a fine mesh. Currently, the users themselves
have to provide such meshes,
e.g., by generating meshes with different cell sizes in external mesh-generation tools.

@@

\todo[inline]{Matthias: Write?}

@@

The 9.5 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_4_2_and_9_5_0.html}
     {around XXX incompatible changes};
\todo{Fix number before release}
see \cite{changes95}. Many of these
incompatibilities change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item
  The class \texttt{hp::DoFHandler} has been removed. The
  \texttt{DoFHandler} class now implements all $hp$-related
  functionality, as it has for the past two releases.

@@

%  \item
  This rework also affects the serialization process of active FE
  indices. You will need to recreate your serialized data \textit{if
    and only if} you work in $hp$-mode, but can continue to use
  previously generated data otherwise. For
  \texttt{parallel::distributed::Triangulation} when used with non-$hp$ data, it is sufficient to increase the version in the metadata file from ``4'' to ``5''.

\item Several old interfaces to \texttt{MatrixFree} have been removed, e.g.,
  initialization functions without \texttt{Mapping} argument and the query
  to the number of cell batches. In
  each case, new interfaces are available.
\end{itemize}
